---
title: "A demo of adaptMT package"
runtime: shiny
output: html_notebook
---

This is a demo of our `adaptMT` package. `adaptMT` has two main components: an API that allows users to specify the working model and algorithms to fit them, as well as a pool of easy-to-use end-to-end wrappers. The former is captured by function `adapt`. The latter includes `adapt_glm`, `adapt_gam` and `adapt_glmnet` in current version (0.1.3.9000) for using GLM, GAM and L1-regularized GLM.

```{r, results='hide'}
# install the "adaptMT" package from github.
# will be submitted to CRAN very soon.
devtools::install_github("lihualei71/adaptMT")
library("adaptMT")
```

We illustrate one of the main function `adapt_glm`, for AdaPT with logistic-Gamma GLM as the working model, on `estrogen` dataset, a gene/drug response dataset from NCBI Gene Expression Omnibus (GEO). `estrogen` dataset consists of gene expression measurements for $n = 22283$ genes, in response to estrogen treatments in breast cancer cells for five groups of patients, with different dosage levels and 5 trials in each. The task is to identify the genes responding to a low dosage. The p-values pi for gene i is obtained by a one-sided permutation test which evaluates evidence for a change in gene expression level between the control group (placebo) and the low-dose group. $\{p_i : i \in [n]\}$ are then ordered according to permutation t-statistics comparing the control and low-dose data, pooled, against data from a higher dosage (with genes that appear to have a strong response at higher dosages placed earlier in the list). The code to compute p-values and the ordering can be found in [Rina Barber's website](http://www.stat.uchicago.edu/~rina/sabha/gene_drug_data_example.R).

In this demo, we subsample the top 5000 genes for illustration.

```{r,results='hide'}
# load the data.
data("estrogen")
# Take the first 5000 genes 
library("dplyr")
estrogen <- select(estrogen, pvals, ord) %>% 
  filter(ord <= 5000)
rownames(estrogen) <- NULL
```

```{r}
head(estrogen, 5)
```

```{r}
summary(estrogen)
```

```{r, echo=FALSE}
renderPlot({
  plot(estrogen$ord, estrogen$pvals, pch = ".", xlab = "order", ylab = "p-values", xaxs = "i", yaxs = "i", main = "Scatter plot of p-values in the (sub-sampled) estrogen dataset")
  }, width = 700, height = 350)
```

Now we execute `adapt_glm` on this dataset. `adapt_glm` assumes a conditional logistic-Gamma GLM as the working model. Specifically, it models the p-values as
$$H_i \mid x_i \sim \pi(x_i), \quad \mathrm{logit}(\pi(x_i))= \phi(x_i)^{T}\beta$$
$$-\log p_i \mid H_i, x_i\sim \left\{\begin{array}{ll} \mathrm{Exp(1)} & H_i = 0\\ \mathrm{Exp(\mu(x))} & H_i = 1\end{array}\right., \quad \frac{1}{\mu(x_i)} = \phi(x_i)^{T}\gamma$$
where $\phi(x)$ is a featurization of $x$. In this example, we use the spline bases, given by `ns` function from `splines` package. For illustration, we choose our candidate models as the above GLMs with $\phi(x)$ being the spline bases with equal-spaced knots and the number of knots ranging from 6-10. We use BIC to select the best model at the initial stage and use the selected model for the following model fitting.

```{r}
# prepare the inputs of AdaPT
# need "splines" package to construct the formula for glm
library("splines")
pvals <- as.numeric(estrogen$pvals)
x <- data.frame(x = as.numeric(estrogen$ord))
formulas <- paste0("ns(x, df = ", 6:10, ")")
formulas
```

`adapt_glm` function provides several user-friendly tools to monitor the progress. For model selection, a progress bar will, by default, be shown in the console that indicates how much proportion of models have been fitted. This can be turned off by setting verbose\$ms = FALSE. Similarly for model fitting, a progress bar can be shown in the console, though not by default, by setting verbose\$fit = TRUE. Also, by default, the progress of the main process will be shown in the console that indicates (1) which target FDR level has been achieved; (2) FDPhat for each target FDR level; (3) number of rejections for each target FDR level.

```{r, warning=FALSE}
# run AdaPT
res <- adapt_glm(x = x, pvals = pvals, pi_formulas = formulas, mu_formulas = formulas)
```

`adapt_1d_plot` gives the plots for the rejection threshold and the estimated local FDR as a function of x (must be univariate without repeated value) for given alpha.

```{r, echo=FALSE}
library("shiny")
sliderInput("alpha", "alpha:", min = 0.01, max = 0.3, value = 0.3, step = 0.01)

renderPlot({
  alpha <- input$alpha
  nrejs <- res$nrejs[floor(alpha * 100)]
  title <- paste0("alpha = ", alpha, ", nrejs = ", nrejs)
  par(mfrow = c(2, 1), mar=c(2, 4.1, 2, 0.15))
  plot_1d_thresh(res, alpha, title, disp_ymax = 1, xlab = "order")
  plot_1d_lfdr(res, alpha, title, disp_ymax = 0.3, xlab = "order")
}, width = 700, height = 650)
```

