---
title: "AdaPT on gene/drug response data"
output: html_notebook
---

This is a demo on running AdaPT, with a logistic-Gamma GLM working model, on the gene/drug response data from NCBI Gene Expression Omnibus (GEO). We start from installing the package and loading the data. The data has been cached in the package.

```{r, results='hide'}
# install the "adaptMT" package from github.
# will be submitted to CRAN very soon.
devtools::install_github("lihualei71/adaptMT")
library("adaptMT")
```

adaptMT provides

```{r}
# load the data and look at the first 10 rows, for brevity
data("estrogen")
```

```{r}
head(estrogen, 10)
```

```{r}
summary(estrogen)
```

Now we , and run it. The function to be used is 

```{r}
# prepare the inputs of AdaPT
# need "splines" package to construct the formula for glm
library("splines")
pvals <- as.numeric(estrogen$pvals)
x <- data.frame(x = as.numeric(estrogen$ord))
pi_formula <- mu_formula <- "ns(x, df = 8)"
```

```{r}
# run AdaPT
res <- adapt_glm(x = x, pvals = pvals, pi_formulas = pi_formula, mu_formulas = mu_formula,
                 verbose = list(print = TRUE, fit = TRUE, ms = TRUE), return_data = FALSE)
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

